---
title: "BST 270: Individual Project"
subtitle: "Reproducible Data Science: How Americans Like Their Steak (FiveThirtyEight)"
author: "Keyao Zhan"
format:
  html: default
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
  gfm: default
code-fold: true
---

## Introduction

The following notebook aims to satisfy the requirements for the individual project component of BST 270: Reproducible Data Science, taken Winter 2025.

#### Motivations and Reproducibility

I aim to reproduce the figure and the table from FiveThirtyEight's [How Americans Like Their Steak](https://fivethirtyeight.com/features/how-americans-like-their-steak/). I will utilize the provided dataset based on a survey testing 550 people about their risk evaluation and steak preference, located at `../data/steak-risk-survey.csv`.

## Setup

First, we load our required packages and required dataset. We utilize the `dplyr`, `knitr` and `ggplot2` library to produce nice figures and tables and process data efficiently.

```{r,echo=F,results='hide'}
library(ggplot2)
library(dplyr)
library(knitr)
```

Here we remove the first two rows which are of no information.

```{r remove}
# Remove first two invalid rows
steak_data <- read.csv("../data/steak-risk-survey.csv",header = T)
steak_data = steak_data[-c(1,2),]
```

The original dataset has too long names and some irrelevant variables, so we extract a new dataset that is useful for our analysis.

```{r}
# Filter and rename the dataset
steak_lottery_data = data.frame(lottery = steak_data$Consider.the.following.hypothetical.situations...br.In.Lottery.A..you.have.a.50..chance.of.success..with.a.payout.of..100...br.In.Lottery.B..you.have.a.90..chance.of.success..with.a.payout.of..20...br..br.Assuming.you.have..10.to.bet..would.you.play.Lottery.A.or.Lottery.B., eat_steak = steak_data$Do.you.eat.steak.,cook_steak = steak_data$How.do.you.like.your.steak.prepared.)
write.csv(steak_lottery_data, file = "../data/steak_lottery_data.csv")
head(steak_lottery_data)
```

## Reproduce the table

The table in the article shows the pertange of steak preparation preference of the people who choose a riskier lottery or a safer lottery.

First we need to remove all the NA (or empty values) in lottery variable, and focus on people who eats steak:

```{r}
df1 = steak_lottery_data %>%
  filter(lottery != "", eat_steak == "Yes")
dim(df1)
```

There are 426 people who answered the lottery question and also eat steak. Next we reproduce the table:

```{r,echo=F}
# Create the counting table
tb1 = table(df1$cook_steak,df1$lottery)
# Rearrange the rows
tb1 = tb1[c(5,3,1,2,4),]

for(i in 1:ncol(tb1)){
  tb1[,i] = tb1[,i]/sum(tb1[,i])
}
colnames(tb1) = c("Riskier lottery","Safer lottery")
```

```{r}
kable(tb1*100,format = "html",digits = 1)
```

**Comment:** We nearly recovered the table in the article, with slight difference.

## Reproduce the figure

The figure shows the percentage of the steak preparation preference of all steak-eating interviewees. First we filter out all people who eat steak.

```{r}
# Create the counting table
df2 = steak_lottery_data %>%
  filter(eat_steak == "Yes")

prep_names = c("Rare","Medium rare","Medium","Medium Well","Well")
df2.1 = data.frame(table(df2$cook_steak)[prep_names]/sum(table(df2$cook_steak))*100)
colnames(df2.1) = c("Cooked","Percentage")
df2.1
```

Next we reproduce the plot in the article:

```{r}
# Reproduce the figure
gg = ggplot(df2.1, aes(x = Cooked, y = Percentage)) +
  geom_bar(stat = "identity", fill = rev(c("#4B2C20", "#754637", "#AA6546", "#C06050", "#C93535"))) +
  coord_flip() +  
  geom_text(aes(label = paste0(round(Percentage), "%")), hjust = -0.2) + 
  labs(
    title = "'How Do You Like Your Steak Prepared?'",
    subtitle = paste("From a survey of", sum(nrow(df2)), "steak-eating responses"),
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10)),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 12),
    axis.ticks = element_blank(),
    aspect.ratio = 0.25
  )
gg
```

**Comment:** From the plot we see that we nearly reproduced the figure, only with slight difference in the percentage.

```{r}
ggsave("../fig/steak_preference.png",plot = gg)
```


## Reflection on the reproducibility

The data is publicly available, and the figure and the table are quite reproducible. I think the only problem arises in the first two rows I removed. As these two rows are indeed invalid ones, I guess the author made some mistakes when storing the data as the csv file.

